Natural variation in teosinte at the domestication locus teosinte branched1 (tb1)

### 2014-05-02 manuscript.tex
new comments, responses.

\jri{Check citation format for AJB. I don't think it's numbers. adjust bibliography style accordingly } \lev{I would rather send to an open access journal. I don't like that no everyone would be able to have access to paper}

@@

Many morphological changes are associated with the domestication of maize, and understanding the genetic basis of these changes has been a focus of maize research for a number of years\citep{Doebley2004}. One of the most dramatic changes is found in plant architecture: domesticated maize is characterized by a central stalk with few tillers and lateral branches terminating in a female inflorescence, while teosinte is highly tillered and bears tassels (male inflorescences) at the end of its lateral branches. The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of organ growth, was identified as a major QTL involved in branching differences between maize and teosinte \citep{Doebley1990, DoebleyStec1991, LukensDoebley1999}. \jri{an odd assortment of cites. why not the 1997 nature paper?} Further work showed that the insertion of a 4.9 kb retrotransposon (Hopscotch) in the upstream control region of \emph{tb1} led to increased expression of this gene, causing the reduction in branching observed in domesticated maize \cite{Studer2011}. The effects of this insertion have been observed in tiller number in maize, but little is known about its role, if any, in teosinte \citep{Studer2011}. Dating of this element has suggested that its insertion predates the domestication of maize, leading to the hypothesis that it was segregating as standing variation in ancient populations of teosinte and increased to high frequency in maize due to selection during domestication \citep{Studer2011}. Furthermore, \citep{StuderDoebley2012} investigated the phenotypic effects of 9 teosinte \emph{tb1} alleles in an isogenic maize background and found that the introgressions sort into three distinct phenotypic classes, suggesting that variation at the \emph{tb1} locus may play a functional role in teosinte. \jri{MBH todo} \lev{hmm?}

@@

In addition to genotyping, we chose a subset of \emph{parviglumis} individuals for sequencing. We chose twelve individuals from each of four populations from Jalisco state, Mexico (San Lorenzo, La Mesa, Ejutla A, and Ejutla B). For amplification and sequencing, we selected two regions approximately 600bp in size from within the 5' UTR of \emph{tb1} (sequenced region 1) and from 1,235 bp upstream of the start of the Hopscotch and 66,169 bp upstream from the start of the \emph{tb1} ORF (sequenced region 2). The 5' UTR (containing sequenced region 1) has been shown to have elevated diversity in ssp. \emph{parviglumis} with respect to maize, while the the area upstream from the Hopscotch (sequenced region 2) has been shown to be critical in determining basal branching and ear architecture \citep{Wang1999, Clark2006}. \jri{is this true? the area upstream of the hopscotch? those papers just found that something upstream was the causal variant, no?} We designed the following primers using PRIMER3 \citep{RozenSkaletsky2000}: for the 5' UTR, {\small 5' GGATAATGTGCACCAGGTGT 3'} and {\small 5' GCGTGCTAGAGACACYTGTTGCT 3'}; for the 50 kb upstream region, {\small 5' TGTCCTCGCCGCAACTC 3'} and {5' TGTACGCCCGCCCCTCATCA 3'} (\ref{FigS1Locus}). We used Taq Polymerase (New England Biolabs) and the following thermal cycler conditions to amplify fragments: 94$^{\circ}$C for 3 min, 30 cycles of 92$^{\circ}$C for 40 s, annealing for 1 min, 72$^{\circ}$C for 40 s, and a final 10 min extension at 72$^{\circ}$C. Annealing temperatures for sequenced region 1 and sequenced region 2 were 59.7$^{\circ}$C and 58.8$^{\circ}$C, respectively. To clean excess primer and dNTPs we added two units of Exonuclease1 and 2.5 units of Antarctic Phosphatase to 8.0 µL of amplification product. This mix was placed on a thermal cycler with the following program: 37$^{\circ}$C for 30 min, 80$^{\circ}$C for 15 min, and a final cool-down step to 4$^{\circ}$C.

@@

\subsection*{Genotyping Analysis} \jri{please drop scripts for bayenv and STRUCTURE into github} \lev{which scripts are you talking about? I used matt's for making structure chunks, and then just command line to run structure, same with BayEnv -- it was Tanjas covariance matrix and then just the basic bayenv commands}

@@

We examined evidence of introgression on chromosome 1 in these same four populations (EjuA, EjuB, MSA, SLO) using STRUCTURE \citep{Falush2003} and the same phased 55K SNP data from \citep{Pyhajarvi2013} that we used for LD analysis, combined with the corresponding SNP data from a diverse panel of 282 maize lines \citep{Cook2012}. SNPs were anchored in a modified version of the IBM genetic map (\citep{Gerke2013}, \url{http://arxiv.org/abs/1307.7313}). We created haplotype blocks using a custom Perl script that grouped SNPs separated by less than 5kb into haplotypes. We ran STRUCTURE at K=2 under the linkage model, performing 3 replicates with an MCMC burn-in of 10,000 steps and 50,000 steps post burn-in. \jri{i'd like this perl script on github, maybe in this repo or as a gist. also structure input file too. all the stuff we'd need to redo this.} \lev{See above note..not all of this or the BayEnv was script'ified}

@@

To investigate the phenotypic effects of the Hopscotch insertion in teosinte, we conducted an initial phenotyping trial (Phenotyping 1). We germinated 250 seeds of \emph{parviglumis} collected in Jalisco state, Mexico (population San Lorenzo) \citep{Hufford2010} where the Hopscotch is segregating at highest frequency (0.44) in our initial genotyping sample set. In order to maximize the likelihood of finding the Hopscotch in our association population we selected seeds from sites where genotyped individuals were homozygous or heterozygous for the insertion. We chose between 10-13 seeds from each of 23 sampling sites. We treated seeds with fungicide and germinated them in petri dishes with filter paper. Successful germinations (206 individuals) were then planted into one gallon size pots with potting soil and randomly spaced one foot apart on greenhouse benches. Plants were watered three times a day with an automatic drip containing 10-20-10 fertilizer. \lev{it ended up being a combination of drip and hand watering because they dried out so much and did better when they had water on the leaves as well as in the soil}

Starting on day 15, we measured tillering index, the ratio of the sum of tiller lengths to the height of the plant \citep{Briggs2007}. Tillering index has been shown to be the most effective way to observe the phenotypic effects of the Hopscotch insertion on plant architecture in maize \citep{Briggs2007}. \jri{was it brgiggs that showed this?} \lev{I believe so, was first to characterize TILL phenotype in backcross} Following initial measurements, we phenotyped plants for tillering index every 5 days through day 40, and then on day 50 and day 60. On day 65 we measured culm diameter between the third and fourth nodes of each plant. Culm diameter is not believed to be correlated with tillering index, or variation at \emph{tb1} (e.g. Hopscotch genotype). Following phenotyping we extracted DNA from all plants using a modified SDS extraction protocol (http://www.ars.usda.gov). We genotyped individuals for the Hopscotch insertion following the protocols listed above. Based on these initial data, we conducted a post hoc power analysis using data from day 40 of phenotyping 1, indicating that a minimum of 71 individuals in each genotype class are needed to detect the observed effect of the Hopscotch on tillering index. \jri{do you still have these posthoc calculations?} \lev{I believe I do, otherwise they are likely in a lab meeting slide on Dropbox -- do you want them in here?}

@@

Resulting tillering index data for each genotype class did not meet the criteria for a repeated measures ANOVA, \jri{why? what criteria?} \lev{basically none of the criteria -- even sample sizes in classes, equal variances, normal distribution, etc.} so we transformed the data using a Box-Cox transformation ($\alpha=0$) implemented in the car package in R \citep{FoxWeisberg2011} to improve the normality and homogeneity of variance among genotype classes.  We analyzed relationships between genotype and tillering index and tiller number using a repeated measures ANOVA through a general linear model function implemented in SAS v.9.3 (SAS Institute Inc., Cary, NC, USA). Additionally, in order to compare any association between Hopscotch genotype and tillering and associations at other presumably unrelated traits, we performed an ANOVA between culm diameter and genotype using the same general linear model in SAS. \jri{please add SAS scripts/code to a gist or something. }

@@

Genotype of the Hopscotch insertion was confirmed with two PCRs for 837 individuals. Among the 247 maize landrace accessions genotyped, all but 8 were homozygous for the presence of the insertion (\ref{TableS1Mays}, \ref{TableS2}). \jri{please fix table/figure references to say table/figure} Within our \emph{parviglumis} and \emph{mexicana} samples we found the Hopscotch insertion segregating in 37 and 4 populations \jri{this is confusing as the map shows <37 populations. we should be consistent about reporting for all pops, or all pops with n>8, etc. or at least be explicit what cutoff we are using for each result}, respectively, and at highest frequency in the states of Jalisco, Colima, and Michoac\'{a}n in central-western Mexico in both subspecies (\ref{Fig1Map}) \jri{the map only shows parviglumis. can we add mexicana? the text makes it seem as if both should be on the map}. We examined Hardy-Weinberg equilibrium in a total of 14 populations (10 \emph{parviglumis} and 4 \emph{mexicana}) with more than 8 individuals sampled per population. Three populations (RIMPA0073, RIMPA0093, and RIMPA0158) show evidence of deviations from expected genotype frequencies under the assumptions of HWE (p<0.05).  \jri{in what direction? too many hets? what's the F? } \lev{they weren't all in the same direction, do you still want me to list out?}

@@

To investigate patterns of sequence diversity and linkage disequilibrium (LD) in the \emph{tb1} region, we sequenced two small ($<$1kb) regions upstream of the tb1 ORF in four populations. After alignment and singleton checking we recovered 40 and 48 segregating sites for the 50kb upstream region and the 5' UTR region, respectively. For region 1, Ejutla A has the highest values of haplotype diversity, and $\theta_\pi$, while Ejutla B and La Mesa have comparable values of these summary statistics, and San Lorenzo has much lower values. Additionally, Tajima's D is strongly negative in the two Ejutla populations and La Mesa, but is more positive in San Lorenzo (\ref{Table2Diversity}). \jri{can drop theta W from table and text. we show pi and D, which is sufficient. fix caption in table. also add Hopscotch allele frequencies to table or list somewhere -- i think it's useful for comparison of seq stats } \lev{the frequencies are in supplemental table 1, do you want them put in this table too, or just me to refer people to supp table 1?}
% Table 2 Diversity Statistics
\begin{table}[htbp]
  \centering
  \caption{Add caption}
    \begin{tabular}{lccccc}
%    \toprule
    Population & \# Haplotypes & Hap. Diversity & $\hat\theta_W$    & $\hat\theta_\pi$    & Tajima's D \\
%    \midrule
    \multicolumn{6}{c}{\textit{Seq. region 2 (50kb upstream)}} \\
    \multicolumn{1}{c}{EJUA} & 8     & 0.89394 & 0.01548 & 0.01763 & 0.6231 \\
    \multicolumn{1}{c}{EJUB} & 8     & 0.89394 & 0.01493 & 0.01591 & 0.29504 \\
    \multicolumn{1}{c}{MSA} & 3     & 0.68182 & 0.01111 & 0.01055 & -0.22212 \\
    \multicolumn{1}{c}{SLO} & 4     & 0.74242 & 0.01167 & 0.01413 & 0.93185 \\
    \multicolumn{6}{c}{\textit{Seq.  region 1(5? UTR)}} \\
    \multicolumn{1}{c}{EJUA} & 8     & 0.85897 & 0.00874 & 0.00527 & -1.64955 \\
    \multicolumn{1}{c}{EJUB} & 5     & 0.70909 & 0.00663 & 0.00378 & -1.83123 \\
    \multicolumn{1}{c}{MSA} & 6     & 0.68182 & 0.00646 & 0.00373 & -1.75506 \\
    \multicolumn{1}{c}{SLO} & 3     & 0.31818 & 0.00176 & 0.00137 & -0.72873 \\
 %   \bottomrule
    \end{tabular}
  \label{Table2Diversity}
\end{table}

@@

To assess the contribution of \emph{tb1} to phenotypic variation in tillering in a natural population, we grew plants from seed sampled from the San Lorenzo population of \emph{parviglumis}, which had a high mean frequency \jri{(X)} of the Hopscotch insertion from our initial genotyping. We measured tillering index (TI), the ratio of the sum of tiller lengths to plant height, for 216 plants from 23 sampling locations within the San Lorenzo population, and genotyped plants for the Hopscotch insertion. \jri{the 23 sampling locales can go in methods and scratch from here}. We found the Hopscotch segregating at a frequency of 0.65 with no significant deviations from expected frequencies under Hardy-Weinberg equilibrium. After performing a repeated measures ANOVA between our transformed tillering index data and Hopscotch genotype we find a weak positive correlation between presence of the Hopscotch and tillering index on day 40 (p=0.0848), but no correlation between tillering index and genotype on any other day (\ref{Fig4Boxplots}). Additionally we find no significant correlation between tiller number and Hopscotch genotype, or culm diameter and Hopscotch genotype in phenotyping 1. \jri{shouldn't we expect a negative correlation betweeh Hop and TI on day 40? need to have an A and B in the figure and explain one is for pheno1 and one is for pheno2. please explain whiskers and dots on figure too.} \lev{sure, I mean, presumably we would expect things with Hop to have a smaller TI}

% Figure 4 Phenotyping BoxPlots
%-------------------------------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   \includegraphics[width=150mm]{Fig4Boxplots.pdf}
    \caption{Box-plot showing tillering index in our greenhouse population from day 20-60. White indicates individuals homozygous for the Hopscotch, light grey represents heterozygotes, and dark grey represents homozygotes for the teosinte (No Hopscotch) allele. Within boxes, dark black lines represent the median, and the edges of the boxes are the first and third quartiles.} 
\label{Fig4Boxplots}
  \end{center}
\end{figure*}
%-------------------------------------------------------------------
\jri{lots of white space in fig 4 and fig. s1 too.} 
We performed a second grow-out  of teosinte to assess whether lighting conditions or sample size may have affected our ability to detect and effect of \emph{tb1}.  For the second grow-out we measured tillering index every five days through day 50 for 302 plants. We found the Hopscotch allele segregating at a frequency of 0.69, \jri{does this violate HWE? didn't we choose numbers of each genotype here, so that HWE isn't appropriate? if so, probably should drop genotype frequency data here} \lev{I don't believe we ended up culling individuals because we had issues with the genotyping for phenotypine 2} with a 0.6 frequency of Hopscotch homozygotes, and a 0.2 frequency of both heterozygotes and homozygotes for the teosinte allele. We found similar patterns, with a weak positive correlation between tillering index and Hopscotch genotype at day 40 (p<0.0611), with no significant correlation on any day. Similarly, relationships between Hopscotch genotype and tiller number, and hopscotch genotype and culm diameter are not significant.

@@

Given that the Hopscotch insertion has been estimated to predate the domestication of maize, it is not surprising that it can be found segregating in populations of teosinte. However, in sampling numerous individuals from many teosinte populations our study provides greater insight into the distribution and prevalence of the Hopscotch in teosinte. While our findings are consistent with a previous study by \citet{Studer2011} in that we identified the Hopscotch allele segregating in teosinte, we find it at higher frequency than previously suggested \citep{Studer2011}. Many of our populations with high frequency of the Hopscotch allele fall in the Jalisco cluster identified by \citet{Fukunaga2005}, possibly suggesting a different history of the \emph{tb1} locus than in the Balsas region where maize was domesticated \citep{Matsuoka2002}. While gene flow from crops into their wild relatives is well-known, (\citep{Ellstrand1999, Zhang2009, Thurber2010, Baack2008, Hubner2012, Wilkes1977, VanHeerwaarden2011, Barrett1983}),\jri{do all these show evidence of introgression AT domestication loci, or just introgression in general?}  \lev{just introgression in general, I believe Thurber and Huff looked at specific domestication loci} our results are more consistent with \citet{Hufford2013} who found resistance to introgression from maize into teosinte \citep{Hufford2013}. Furthermore, \citet{Hufford2013} showed that domestication loci, such as \emph{tb1}, are particularly resistant to introgression in both directions of gene flow (i.e., maize to teosinte and teosinte to maize) \citep{Hufford2013}.

@@

The phenotypic effects of the Hopscotch insertion in domesticated maize have been well documented \citep{Clark2006, Studer2011}, \jri{did clark do phenotyping?} \lev{yes they phenotyped their bc1 i believe, 8 traits shown to correlate with tb1 variation} and \citet{Weber2007} have described its effects in partially inbred lines of teosinte \citep{Weber2007} \jri{i don't think these were inbred.please doublecheck.}. Our study is the first to explicitly examine the phenotypic effects of the Hopscotch insertion in individuals sampled from a natural population of teosinte. \jri{isn't this what weber did?? for 70+ populations!?} However, we found no significant effect of the Hopscotch on tillering index or tiller number in our phenotyping experiments, and the effect of the Hopscotch insertion in teosinte is discordant with that of maize. The lack of correlation between Hopscotch genotype and tillering index or tiller number is surprising given its effects in maize. It is certainly possible that even though previous data demonstrate an effect of the Hopscotch on tillering in maize \citep{Studer2011}, that the effect of the Hopscotch in teosinte is more complicated and may be more difficult to observe. Moreover, \emph{tb1} is a single gene in a complex pathway that affects branching and tillering traits, and perhaps in combination with alleles at other loci the phenotypic effects of the Hopscotch on tillering may not be consistent. \jri{thsi section still needs work. i think we have to do more here. weber shows an association between SNPs in tb1 and branch length. we need to discuss that!}

>>>>

\jri{Check citation format for AJB. I don't think it's numbers. adjust bibliography style accordingly } \lev{I would rather send to an open access journal. I don't like that no everyone would be able to have access to paper} \jri{AJB allows open access for $500 which is cheaper than PLoS ONE}

@@

Many morphological changes are associated with the domestication of maize, and understanding the genetic basis of these changes has been a focus of maize research for a number of years\citep{Doebley2004}. One of the most dramatic changes is found in plant architecture: domesticated maize is characterized by a central stalk with few tillers and lateral branches terminating in a female inflorescence, while teosinte is highly tillered and bears tassels (male inflorescences) at the end of its lateral branches. The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of organ growth, was identified as a major QTL involved in branching differences between maize and teosinte \citep{Doebley1990, DoebleyStec1991, LukensDoebley1999}. \jri{an odd assortment of cites. why not the 1997 nature paper?} Further work showed that the insertion of a 4.9 kb retrotransposon (Hopscotch) in the upstream control region of \emph{tb1} led to increased expression of this gene, causing the reduction in branching observed in domesticated maize \cite{Studer2011}. The effects of this insertion have been observed in tiller number in maize, but little is known about its role, if any, in teosinte \citep{Studer2011}. Dating of this element has suggested that its insertion predates the domestication of maize, leading to the hypothesis that it was segregating as standing variation in ancient populations of teosinte and increased to high frequency in maize due to selection during domestication \citep{Studer2011}. Furthermore, \citep{StuderDoebley2012} investigated the phenotypic effects of 9 teosinte \emph{tb1} alleles in an isogenic maize background and found that the introgressions sort into three distinct phenotypic classes, suggesting that variation at the \emph{tb1} locus may play a functional role in teosinte.

@@

In addition to genotyping, we chose a subset of \emph{parviglumis} individuals for sequencing. We chose twelve individuals from each of four populations from Jalisco state, Mexico (San Lorenzo, La Mesa, Ejutla A, and Ejutla B). For amplification and sequencing, we selected two regions approximately 600bp in size from within the 5' UTR of \emph{tb1} (sequenced region 1) and from 1,235 bp upstream of the start of the Hopscotch and 66,169 bp upstream from the start of the \emph{tb1} ORF (sequenced region 2). The 5' UTR (containing sequenced region 1) has been shown to have elevated diversity in ssp. \emph{parviglumis} with respect to maize, while the the area upstream from the Hopscotch (sequenced region 2) has been shown to be critical in determining basal branching and ear architecture \citep{Wang1999, Clark2006}. \jri{is this true? the area upstream of the hopscotch? those papers just found that something upstream was the causal variant, no?} We designed the following primers using PRIMER3 \citep{RozenSkaletsky2000}: for the 5' UTR, {\small 5' GGATAATGTGCACCAGGTGT 3'} and {\small 5' GCGTGCTAGAGACACYTGTTGCT 3'}; for the 50 kb upstream region, {\small 5' TGTCCTCGCCGCAACTC 3'} and {5' TGTACGCCCGCCCCTCATCA 3'} (\ref{FigS1Locus}). We used Taq Polymerase (New England Biolabs) and the following thermal cycler conditions to amplify fragments: 94$^{\circ}$C for 3 min, 30 cycles of 92$^{\circ}$C for 40 s, annealing for 1 min, 72$^{\circ}$C for 40 s, and a final 10 min extension at 72$^{\circ}$C. Annealing temperatures for sequenced region 1 and sequenced region 2 were 59.7$^{\circ}$C and 58.8$^{\circ}$C, respectively. To clean excess primer and dNTPs we added two units of Exonuclease1 and 2.5 units of Antarctic Phosphatase to 8.0 $\mu$ L of amplification product. This mix was placed on a thermal cycler with the following program: 37$^{\circ}$C for 30 min, 80$^{\circ}$C for 15 min, and a final cool-down step to 4$^{\circ}$C.

@@

\subsection*{Genotyping Analysis} \jri{please drop scripts for bayenv and STRUCTURE into github} \lev{which scripts are you talking about? I used matt's for making structure chunks, and then just command line to run structure, same with BayEnv -- it was Tanjas covariance matrix and then just the basic bayenv commands} \jri{matt's scritp for structure should be up there. and even a readme documenting the commandline for bayenv and including the covariance matrix. basically whatever someone would need to be able to literally repeat what you did.}

@@

We examined evidence of introgression on chromosome 1 in these same four populations (EjuA, EjuB, MSA, SLO) using STRUCTURE \citep{Falush2003} and the same phased 55K SNP data from \citep{Pyhajarvi2013} that we used for LD analysis, combined with the corresponding SNP data from a diverse panel of 282 maize lines \citep{Cook2012}. SNPs were anchored in a modified version of the IBM genetic map (\citep{Gerke2013}, \url{http://arxiv.org/abs/1307.7313}). We created haplotype blocks using a custom Perl script that grouped SNPs separated by less than 5kb into haplotypes. We ran STRUCTURE at K=2 under the linkage model, performing 3 replicates with an MCMC burn-in of 10,000 steps and 50,000 steps post burn-in. \jri{i'd like this perl script on github, maybe in this repo or as a gist. also structure input file too. all the stuff we'd need to redo this.} \lev{See above note..not all of this or the BayEnv was script'ified} \jri{sure, even command line info should be included where possible.  idea is to maximize reproducibility -- either other people or subsequent students. for example, matt has a student who wants to work on tb1 in natural pops, and she might want to try/redo some of these analyses with the same or new data.}

@@

To investigate the phenotypic effects of the Hopscotch insertion in teosinte, we conducted an initial phenotyping trial (Phenotyping 1). We germinated 250 seeds of \emph{parviglumis} collected in Jalisco state, Mexico (population San Lorenzo) \citep{Hufford2010} where the Hopscotch is segregating at highest frequency (0.44) in our initial genotyping sample set. In order to maximize the likelihood of finding the Hopscotch in our association population we selected seeds from sites where genotyped individuals were homozygous or heterozygous for the insertion. We chose between 10-13 seeds from each of 23 sampling sites. We treated seeds with fungicide and germinated them in petri dishes with filter paper. Successful germinations (206 individuals) were then planted into one gallon size pots with potting soil and randomly spaced one foot apart on greenhouse benches. Plants were watered three times a day.

Starting on day 15, we measured tillering index, the ratio of the sum of tiller lengths to the height of the plant \citep{Briggs2007}. Tillering index has been shown to be the most effective way to observe the phenotypic effects of the Hopscotch insertion on plant architecture in maize \citep{Briggs2007}. \jri{was it brgiggs that showed this?} \lev{I believe so, was first to characterize TILL phenotype in backcross} \jri{I think this cite should probably be one of the earlier Doebley papers mapping to tb1 QTL, as those showed the QTL had effect on plant architecture, no?} Following initial measurements, we phenotyped plants for tillering index every 5 days through day 40, and then on day 50 and day 60. On day 65 we measured culm diameter between the third and fourth nodes of each plant. Culm diameter is not believed to be correlated with tillering index, or variation at \emph{tb1} (e.g. Hopscotch genotype). Following phenotyping we extracted DNA from all plants using a modified SDS extraction protocol (http://www.ars.usda.gov). We genotyped individuals for the Hopscotch insertion following the protocols listed above. Based on these initial data, we conducted a post hoc power analysis using data from day 40 of phenotyping 1, indicating that a minimum of 71 individuals in each genotype class are needed to detect the observed effect of the Hopscotch on tillering index. \jri{do you still have these posthoc calculations?} \lev{I believe I do, otherwise they are likely in a lab meeting slide on Dropbox -- do you want them in here?} \jri{would be good to include in a github, again so we could go back and reassess how we do things in case, for example, we decide to give the greenhouse experiment a 3rd try (yes, i'm a masochist)}

@@

Resulting tillering index data for each genotype class did not meet the criteria for a repeated measures ANOVA, so we transformed the data using a Box-Cox transformation ($\alpha=0$) implemented in the car package in R \citep{FoxWeisberg2011} to improve the normality and homogeneity of variance among genotype classes.  We analyzed relationships between genotype and tillering index and tiller number using a repeated measures ANOVA through a general linear model function implemented in SAS v.9.3 (SAS Institute Inc., Cary, NC, USA). Additionally, in order to compare any association between Hopscotch genotype and tillering and associations at other presumably unrelated traits, we performed an ANOVA between culm diameter and genotype using the same general linear model in SAS. \jri{please add SAS scripts/code to a gist or something. }

@@

Genotype of the Hopscotch insertion was confirmed with two PCRs for 837 individuals. Among the 247 maize landrace accessions genotyped, all but 8 were homozygous for the presence of the insertion (\ref{TableS1Mays}, \ref{TableS2}). \jri{please fix table/figure references to say table/figure} Within our \emph{parviglumis} and \emph{mexicana} samples we found the Hopscotch insertion segregating in 37 and 4 populations \jri{this is confusing as the map shows <37 populations. we should be consistent about reporting for all pops, or all pops with n>8, etc. or at least be explicit what cutoff we are using for each result}, respectively, and at highest frequency in the states of Jalisco, Colima, and Michoac\'{a}n in central-western Mexico in both subspecies (\ref{Fig1Map}) \jri{the map only shows parviglumis. can we add mexicana? the text makes it seem as if both should be on the map}. We examined Hardy-Weinberg equilibrium in a total of 14 populations (10 \emph{parviglumis} and 4 \emph{mexicana}) with more than 8 individuals sampled per population. Three populations (RIMPA0073, RIMPA0093, and RIMPA0158) show evidence of deviations from expected genotype frequencies under the assumptions of HWE (p<0.05).  \jri{in what direction? too many hets? what's the F? } \lev{they weren't all in the same direction, do you still want me to list out?} \jri{no need to list all, but if there were sig. deviations in multiple directions in different pops, that is worth saying. maybe could list them all out in supp. table?}

@@

To investigate patterns of sequence diversity and linkage disequilibrium (LD) in the \emph{tb1} region, we sequenced two small ($<$1kb) regions upstream of the tb1 ORF in four populations. After alignment and singleton checking we recovered 40 and 48 segregating sites for the 50kb upstream region and the 5' UTR region, respectively. For region 1, Ejutla A has the highest values of haplotype diversity, and $\theta_\pi$, while Ejutla B and La Mesa have comparable values of these summary statistics, and San Lorenzo has much lower values. Additionally, Tajima's D is strongly negative in the two Ejutla populations and La Mesa, but is more positive in San Lorenzo (\ref{Table2Diversity}). \jri{can drop theta W from table and text. we show pi and D, which is sufficient. fix caption in table. also add Hopscotch allele frequencies to table or list somewhere -- i think it's useful for comparison of seq stats } \lev{the frequencies are in supplemental table 1, do you want them put in this table too, or just me to refer people to supp table 1?} \jri{yeah refer to table is fine}
% Table 2 Diversity Statistics
\begin{table}[htbp]
  \centering
  \caption{Add caption}
    \begin{tabular}{lccccc}
%    \toprule
    Population & \# Haplotypes & Hap. Diversity & $\hat\theta_W$    & $\hat\theta_\pi$    & Tajima's D \\
%    \midrule
    \multicolumn{6}{c}{\textit{Seq. region 2 (50kb upstream)}} \\
    \multicolumn{1}{c}{EJUA} & 8     & 0.89394 & 0.01548 & 0.01763 & 0.6231 \\
    \multicolumn{1}{c}{EJUB} & 8     & 0.89394 & 0.01493 & 0.01591 & 0.29504 \\
    \multicolumn{1}{c}{MSA} & 3     & 0.68182 & 0.01111 & 0.01055 & -0.22212 \\
    \multicolumn{1}{c}{SLO} & 4     & 0.74242 & 0.01167 & 0.01413 & 0.93185 \\
    \multicolumn{6}{c}{\textit{Seq.  region 1(5? UTR)}} \\
    \multicolumn{1}{c}{EJUA} & 8     & 0.85897 & 0.00874 & 0.00527 & -1.64955 \\
    \multicolumn{1}{c}{EJUB} & 5     & 0.70909 & 0.00663 & 0.00378 & -1.83123 \\
    \multicolumn{1}{c}{MSA} & 6     & 0.68182 & 0.00646 & 0.00373 & -1.75506 \\
    \multicolumn{1}{c}{SLO} & 3     & 0.31818 & 0.00176 & 0.00137 & -0.72873 \\
 %   \bottomrule
    \end{tabular}
  \label{Table2Diversity}
\end{table}

@@

To assess the contribution of \emph{tb1} to phenotypic variation in tillering in a natural population, we grew plants from seed sampled from the San Lorenzo population of \emph{parviglumis}, which had a high mean frequency \jri{(X)} of the Hopscotch insertion from our initial genotyping. We measured tillering index (TI), the ratio of the sum of tiller lengths to plant height, for 216 plants from 23 sampling locations within the San Lorenzo population, and genotyped plants for the Hopscotch insertion. \jri{the 23 sampling locales can go in methods and scratch from here}. We found the Hopscotch segregating at a frequency of 0.65 with no significant deviations from expected frequencies under Hardy-Weinberg equilibrium. After performing a repeated measures ANOVA between our transformed tillering index data and Hopscotch genotype we find a weak positive correlation between presence of the Hopscotch and tillering index on day 40 (p=0.0848), but no correlation between tillering index and genotype on any other day (\ref{Fig4Boxplots}). Additionally we find no significant correlation between tiller number and Hopscotch genotype, or culm diameter and Hopscotch genotype in phenotyping 1. \jri{shouldn't we expect a negative correlation betweeh Hop and TI on day 40? need to have an A and B in the figure and explain one is for pheno1 and one is for pheno2. please explain whiskers and dots on figure too.} \lev{sure, I mean, presumably we would expect things with Hop to have a smaller TI} \jri{yup, but we should mention that the expectation is negative/}

% Figure 4 Phenotyping BoxPlots
%-------------------------------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   \includegraphics[width=150mm]{Fig4Boxplots.pdf}
    \caption{Box-plot showing tillering index in our greenhouse population from day 20-60. White indicates individuals homozygous for the Hopscotch, light grey represents heterozygotes, and dark grey represents homozygotes for the teosinte (No Hopscotch) allele. Within boxes, dark black lines represent the median, and the edges of the boxes are the first and third quartiles.} 
\label{Fig4Boxplots}
  \end{center}
\end{figure*}
%-------------------------------------------------------------------
\jri{lots of white space in fig 4 and fig. s1 too.} 
We performed a second grow-out  of teosinte to assess whether lighting conditions or sample size may have affected our ability to detect and effect of \emph{tb1}.  For the second grow-out we measured tillering index every five days through day 50 for 302 plants. We found the Hopscotch allele segregating at a frequency of 0.69, \jri{is it in HWE in this pop?} with a 0.6 frequency of Hopscotch homozygotes, and a 0.2 frequency of both heterozygotes and homozygotes for the teosinte allele. We found similar patterns, with a weak positive correlation between tillering index and Hopscotch genotype at day 40 (p<0.0611), with no significant correlation on any day. Similarly, relationships between Hopscotch genotype and tiller number, and hopscotch genotype and culm diameter are not significant.

@@

Given that the Hopscotch insertion has been estimated to predate the domestication of maize, it is not surprising that it can be found segregating in populations of teosinte. However, in sampling numerous individuals from many teosinte populations our study provides greater insight into the distribution and prevalence of the Hopscotch in teosinte. While our findings are consistent with a previous study by \citet{Studer2011} in that we identified the Hopscotch allele segregating in teosinte, we find it at higher frequency than previously suggested \citep{Studer2011}. Many of our populations with high frequency of the Hopscotch allele fall in the Jalisco cluster identified by \citet{Fukunaga2005}, possibly suggesting a different history of the \emph{tb1} locus than in the Balsas region where maize was domesticated \citep{Matsuoka2002}. While gene flow from crops into their wild relatives is well-known, (\citep{Ellstrand1999, Zhang2009, Thurber2010, Baack2008, Hubner2012, Wilkes1977, VanHeerwaarden2011, Barrett1983}), our results are more consistent with \citet{Hufford2013} who found resistance to introgression from maize into teosinte \citep{Hufford2013}. Furthermore, \citet{Hufford2013} showed that domestication loci, such as \emph{tb1}, are particularly resistant to introgression in both directions of gene flow (i.e., maize to teosinte and teosinte to maize) \citep{Hufford2013}.

@@

The phenotypic effects of the Hopscotch insertion in domesticated maize have been well documented \citep{Clark2006, Studer2011}, and \citet{Weber2007} have described its effects in partially inbred lines of teosinte \citep{Weber2007} \jri{i don't think these were inbred.please doublecheck.}. Our study is the first to explicitly examine the phenotypic effects of the Hopscotch insertion in individuals sampled from a natural population of teosinte. \jri{isn't this what weber did?? for 70+ populations!?} However, we found no significant effect of the Hopscotch on tillering index or tiller number in our phenotyping experiments, and the effect of the Hopscotch insertion in teosinte is discordant with that of maize. The lack of correlation between Hopscotch genotype and tillering index or tiller number is surprising given its effects in maize. It is certainly possible that even though previous data demonstrate an effect of the Hopscotch on tillering in maize \citep{Studer2011}, that the effect of the Hopscotch in teosinte is more complicated and may be more difficult to observe. Moreover, \emph{tb1} is a single gene in a complex pathway that affects branching and tillering traits, and perhaps in combination with alleles at other loci the phenotypic effects of the Hopscotch on tillering may not be consistent. \jri{thsi section still needs work. i think we have to do more here. weber shows an association between SNPs in tb1 and branch length. we need to discuss that!}
